Staged Abstract Interpreters

### 2019-04-05 paper/10_intro.tex
paper

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. We observe at most 14x times faster than the unstage abstract
interpreter.
However, we do not lose any advantages of this high-level programming style
(e.g., modularity and extensibility); the meaning of the program also remains
the same, particularly, the soundness of static analysis after staging.

@@

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

\todo{mention LMS somewhere}

>>>>

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe at most 14x times faster
than the unstage abstract interpreter. However, we do not lose any advantages of
this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of
static analysis after staging.

@@

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

### 2019-04-05 paper/20_prelim.tex
paper

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily. In Section~\ref{cases_imp}, we
will add more imperative features to the language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguise any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use data-types to represent programs. This design choice allows us to
easily use different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

>>>>

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily.
%In Section~\ref{cases_imp}, we will add more imperative features to the
language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguise any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use algebraic data-types to represent programs. This design choice allows us to
easily use different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

### 2019-04-05 paper/30_stageconc.tex
paper

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to let them works on staged data.
As we mentioned before, we elaborate how to handle functions and applications in
detail. One of the desired goal is to eliminate the monadic layers through
staging, this becomes a little bit tricky for clsoures, because we need actually
compiling the lambda term instead of generating a @CloV@ value, but the @ev(e)@
returns a monad and prevents the specialization. To make this happen, when
closing a lambda term, we need to staging into the body expression by
\textit{collpasing} the @AnsM@ monads to normal values. The
following function @close@ illustrates the idea: @f@ is a current stage function
but takes two next-stage values @v@ and $\sigma$, inside of which, we eagerly
collapse the monads to values by @ev(e)(ρ_*)(σ_*)@, which has type
@Rep[(Value, Store)]@. Then we generate an intermediate representation of the compiled
closure using @emit_compiled_clo@. The intermediate representation stands for a
next-stage value, and contains the compiled function, which will be generated as
a next-stage function.

@@

When applying a function in @ap_clo@, we generate an intermediate representation
for closure application through @emit_ap_clo@, assuming that the @rator@ is already an
IR of compiled closure, @rand@ is an arbitrary stage value, and $\sigma$ is the
latest store. @emit_ap_clo@ represents a next-stage value of type
@Rep[(Value, Store)]@, which is a collapsed monadic value from next stage, now
we need to reify the value and store back into the monad stack through
@put_store@ and @yield@.

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces IR
node in LMS. We define these IR node as follow: \note{Exp, Def, quote are omitted}

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[((Value,Store)) => (Value, Store)]) extends Rep[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Rep[(Value, Store)]
\end{lstlisting}

@@

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar$rand, $\textdollar$σ)")
    case IRCompiledClo(f) => emitValDef(sym, s"CompiledClo($\textdollar$f)")
    ...
  }
\end{lstlisting}

>>>>

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to let them works on staged data.
As we mentioned before, we elaborate how to handle functions and applications in
detail. One of the desired goal is to eliminate the monadic layers through
staging, this becomes a little bit tricky for clsoures, because we need actually
compiling the lambda term instead of generating a @CloV@ value, but the @ev(e)@
returns a monad and prevents the specialization. To make this happen, when
closing a lambda term, we need to staging into the body expression by
\textit{collpasing} the @AnsM@ monads to normal values. The
following function @close@ illustrates the idea: @f@ is a current stage function
but takes two next-stage values @v@ and $\sigma$, inside of which, we eagerly
collapse the monads to values by @ev(e)(ρ_*)(σ_*)@, which has type
@Rep[(Value, Store)]@. Then we generate the next-stage value of the compiled
closure using @emit_compiled_clo@, containing the compiled closure.

@@

When applying a function in @ap_clo@, we also generate the next-stage value
for closure application through @emit_ap_clo@, assuming that the @rator@ is already an
IR of compiled closure, @rand@ is an arbitrary stage value, and $\sigma$ is the
latest store. @emit_ap_clo@ represents the result value of application, which has type
@Rep[(Value, Store)]@. Note that @emit_ap_clo@ is a ordinary value from the
future, and there is no monads in the future, so we need to reify the value and
store back into the monad stack through @put_store@ and @yield@.

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces
intermediate representations for the denoted operations in LMS. These IR are
wrappers of the underlying staged data. We define these IR nodes using
@case class@ extending from @Def[T]@, meaning that they are next-staged definitions of
type @T@.

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[((Value,Store)) => (Value, Store)], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[(Value, Store)]
\end{lstlisting}

@@

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar$rand, $\textdollar$σ)")
    case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar$f, $\textdollar$λ, $\textdollar$ρ)")
    ...
  }
\end{lstlisting}

### 2019-04-05 paper/50_stageabs.tex
paper

Since our monad stack is uniformly using staged data, so the staged @SetT@ now
stores a staged value of type @Rep[Set[A]]@, inside of a staged monad @M@.
\todo{staged SetT implementation}

@@

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represents the result of application, and finally we
put the @out@ cache, store and values back into the monads. The @emit_ap_clo@
method takes the target closure @clo@, argument @rand@, store and caches.

@@

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value, Store)], Cache)]) extends Rep[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Rep[(Set[(Value, Store)], Cache)]
\end{lstlisting}

>>>>

Since our monad stack is uniformly using staged data, so the staged @SetT@ now
stores a staged value of type @Rep[Set[A]]@, inside of anohter staged monad @M@.

@@

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represents the result of application from future.
Finally we put the future @out@ cache, store and values back into the
current-stage monads. The @emit_ap_clo@ method takes the target closure @clo@,
argument @rand@, store and caches.

@@

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value, Store)], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

### 2019-04-05 paper/60_cases.tex
paper

\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield KCFAAddr(x, τ)
\end{lstlisting}

>>>>

\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we unit to turn a current-stage constant to next-stage value
\end{lstlisting}

### 2019-04-05 paper/70_eval.tex
paper

\begin{figure}[h]
\tiny
\begin{tabular}{@{}lllll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ \\ \midrule
    @rsa@               & 451  & 0.308      & 0.064      & 459.15\%          \\
    @church@            & 120  & 71.811     & 23.990     & 299.34\%          \\
    @fermat@            & 310  & 0.043      & 0.006      & 716.67\%          \\
    @mbrotZ@            & 331  & 1.112      & 0.214      & 519.63\%          \\
    @lattice@           & 609  & 31.874     & 8.816      & 358.69\%          \\
    @kcfa-worst-16@     & 182  & 0.083      & 0.013      & 638.46\%          \\
    @kcfa-worst-32@     & 358  & 0.452      & 0.044      & 1027.27\%         \\
    @kcfa-worst-64@     & 774  & 3.329      & 0.236      & 1410.59\%         \\
    @solovay-strassen@  & 523  & 5.94       & 1.189      & 428.43\%          \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

>>>>

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}lllll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ \\ \midrule
    @fib@               & 32   & 0.013      & 0.002      & 650.00\%          \\
    @rsa@               & 451  & 0.042      & 0.232      & 552.38\%          \\
    @church@            & 120  & 71.811     & 23.990     & 299.34\%          \\
    @fermat@            & 310  & 0.043      & 0.006      & 716.67\%          \\
    @mbrotZ@            & 331  & 1.112      & 0.214      & 519.63\%          \\
    @lattice@           & 609  & 31.874     & 8.816      & 358.69\%          \\
    @kcfa-worst-16@     & 182  & 0.083      & 0.013      & 638.46\%          \\
    @kcfa-worst-32@     & 358  & 0.452      & 0.044      & 1027.27\%         \\
    @kcfa-worst-64@     & 774  & 3.329      & 0.236      & 1410.59\%         \\
    @solovay-strassen@  & 523  & 5.94       & 1.189      & 428.43\%          \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

### 2019-04-05 paper/abstract.tex
paper

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the applicability on various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of \note{10}x with staging for a value-flow analysis.

>>>>

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the applicability on various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 6.6x with staging for control-flow analysis.
